---
title: "Preregistered Outlier Diagnostics (Supplementary)"
subtitle: "CPX-25-0117 — Written Exposure Therapy for Transdiagnostic Anxiety"
author: "Elad Zlotnick, Jonathan D. Huppert"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: pdflatex
    keep_tex: false
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{pdflscape}
  - \DeclareUnicodeCharacter{2014}{\textemdash}
  - \DeclareUnicodeCharacter{00D7}{$\times$}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  cache   = TRUE
)

library(tidyverse)
library(rethinking)
library(here)
library(knitr)
library(kableExtra)

source(here('scripts', 'load_data.R'))
source(here('scripts', 'helpers.R'))
source(here('scripts', 'model.growth.R'))
source(here('scripts', 'model.change.R'))
source(here('scripts', 'model.mechanisms.R'))
```

```{r diagnose-fn, include=FALSE}
diagnose_model <- function(m, d, model_label) {
  w <- WAIC(m, pointwise = TRUE)
  p <- PSIS(m, pointwise = TRUE)
  penalty_col <- intersect(c('penalty', 'p_waic', 'pWAIC'), names(w))[1]
  k_col       <- intersect(c('k', 'pareto_k'),              names(p))[1]
  stopifnot(!is.na(penalty_col), !is.na(k_col))
  if (nrow(w) != nrow(d) || nrow(p) != nrow(d)) {
    warning(sprintf("[%s] Row mismatch. Skipped.", model_label))
    return(NULL)
  }
  d %>%
    mutate(
      model     = model_label,
      row_id    = row_number(),
      p_waic    = w[[penalty_col]],
      pareto_k  = p[[k_col]],
      flag_waic = p_waic   > 1,
      flag_psis = pareto_k > .5,
      flag_both = flag_waic & flag_psis,
      flag_psis_severe = pareto_k > .7
    )
}

summarize_diag <- function(diag) {
  if (is.null(diag)) return(NULL)
  diag %>%
    summarize(
      n_obs            = n(),
      n_participants   = n_distinct(id),
      max_p_waic       = max(p_waic, na.rm = TRUE),
      max_pareto_k     = max(pareto_k, na.rm = TRUE),
      n_waic_gt1       = sum(flag_waic, na.rm = TRUE),
      n_k_gt.5         = sum(flag_psis, na.rm = TRUE),
      n_k_gt.7         = sum(flag_psis_severe, na.rm = TRUE),
      n_both           = sum(flag_both, na.rm = TRUE),
      pct_both         = round(mean(flag_both, na.rm = TRUE) * 100, 3)
    )
}

diagnose_mechanism <- function(m, d_x, d_m, label_x, label_m = NULL) {
  w <- WAIC(m, pointwise = TRUE)
  p <- PSIS(m, pointwise = TRUE)
  penalty_col <- intersect(c('penalty', 'p_waic', 'pWAIC'), names(w))[1]
  k_col       <- intersect(c('k', 'pareto_k'),              names(p))[1]
  stopifnot(!is.na(penalty_col), !is.na(k_col))
  n_x <- nrow(d_x); n_m <- nrow(d_m); n_tot <- n_x + n_m
  if (nrow(w) == n_tot && nrow(p) == n_tot) {
    diag_x <- d_x %>%
      mutate(model = label_x, row_id = row_number(),
             p_waic = w[[penalty_col]][1:n_x],
             pareto_k = p[[k_col]][1:n_x],
             flag_waic = p_waic > 1, flag_psis = pareto_k > .5,
             flag_both = flag_waic & flag_psis,
             flag_psis_severe = pareto_k > .7)
    diag_m <- d_m %>%
      mutate(model = label_m %||% paste0(label_x, '.m'), row_id = row_number(),
             p_waic = w[[penalty_col]][(n_x + 1):n_tot],
             pareto_k = p[[k_col]][(n_x + 1):n_tot],
             flag_waic = p_waic > 1, flag_psis = pareto_k > .5,
             flag_both = flag_waic & flag_psis,
             flag_psis_severe = pareto_k > .7)
    return(bind_rows(diag_x, diag_m))
  }
  if (nrow(w) == n_x && nrow(p) == n_x) {
    return(d_x %>%
      mutate(model = label_x, row_id = row_number(),
             p_waic = w[[penalty_col]], pareto_k = p[[k_col]],
             flag_waic = p_waic > 1, flag_psis = pareto_k > .5,
             flag_both = flag_waic & flag_psis,
             flag_psis_severe = pareto_k > .7))
  }
  warning(sprintf("[%s] Unexpected row mismatch. Skipped.", label_x))
  NULL
}
```

```{r growth-diag, include=FALSE}
growth_model_names <- list(
  oasis      = 'oasis',
  wsas       = 'wsas',
  engagement = 'enagement'   # preserve original typo in the cached model name
)

growth_diag <- map_dfr(names(growth_model_names), function(mname) {
  d <- data_long %>%
    filter(measure == mname, !is.na(value), wave > 0) %>%
    left_join(demographics.data, by = 'id')
  m <- model.growth(d, growth_model_names[[mname]], force = FALSE)
  diagnose_model(m, d, paste0('growth.', mname))
})
```

```{r change-diag, include=FALSE}
change_measures <- c(
  'dts', 'scs', 'ifs',
  'taf_m', 'taf_l',
  'bcss_self_neg', 'bcss_other_neg', 'bcss_self_pos', 'bcss_other_pos',
  'mcq_confidence', 'mcq_worry', 'mcq_self', 'mcq_danger', 'mcq_control'
)

change_diag <- map_dfr(change_measures, function(mname) {
  d <- data_long %>%
    filter(measure == mname, !is.na(value), condition != 'waitlist', wave > 0)
  m <- model.change(d, model_name = mname, force = FALSE)
  diagnose_model(m, d, paste0('change.', mname))
})
```

```{r mechanism-diag, include=FALSE}
oasis_treatment <- data_long %>%
  filter(measure == 'oasis', !is.na(value), condition == 'treatment', wave > 0) %>%
  left_join(demographics.data, by = 'id')

oasis_control <- data_long %>%
  filter(measure == 'oasis', !is.na(value), condition == 'control', wave > 0) %>%
  left_join(demographics.data, by = 'id')

mech_treatment_measures <- c('dts', 'bcss_self_pos', 'ifs', 'tiii',
                             'mcq_confidence', 'mcq_danger')

mech_diag_1 <- map_dfr(mech_treatment_measures, function(mname) {
  d_m <- data_long %>%
    filter(measure == mname, !is.na(value), condition == 'treatment', wave > 0)
  m <- model.mechanism(oasis_treatment, d_m, model_name = mname, force = FALSE)
  diagnose_mechanism(m, oasis_treatment, d_m,
                     label_x = paste0('mechanism.', mname, '.oasis'))
})

bsam_wet <- data_long %>%
  filter(measure == 'bsam.pre', condition == 'treatment',
         wave %in% 1:2, !is.na(value)) %>%
  mutate(wave = if_else(wave == 1, 1, 4))

mech_bsam_wet <- model.mechanism(oasis_treatment, bsam_wet,
                                 model_name = 'bsam.change.all', force = FALSE)
mech_diag_bsam_wet <- diagnose_mechanism(
  mech_bsam_wet, oasis_treatment, bsam_wet,
  label_x = 'mechanism.bsam.change.all.oasis')

bsam_ew <- data_long %>%
  filter(measure == 'bsam.pre', condition == 'control',
         wave %in% 1:2, !is.na(value)) %>%
  mutate(wave = if_else(wave == 1, 1, 4))

mech_bsam_ew <- model.mechanism(oasis_control, bsam_ew,
                                model_name = 'bsam.change.all.ew', force = FALSE)
mech_diag_bsam_ew <- diagnose_mechanism(
  mech_bsam_ew, oasis_control, bsam_ew,
  label_x = 'mechanism.bsam.change.all.ew.oasis')

dts_ew <- data_long %>%
  filter(measure == 'dts', !is.na(value), condition == 'control', wave > 0)

mech_dts_ew <- model.mechanism(oasis_control, dts_ew,
                               model_name = 'dts.ew', force = FALSE)
mech_diag_dts_ew <- diagnose_mechanism(
  mech_dts_ew, oasis_control, dts_ew,
  label_x = 'mechanism.dts.ew.oasis')

dts_treatment <- data_long %>%
  filter(measure == 'dts', !is.na(value), condition == 'treatment', wave > 0) %>%
  left_join(demographics.data, by = 'id')

mech_dts_bsam <- model.mechanism(dts_treatment, bsam_wet,
                                 model_name = 'dts.bsam.wet', force = FALSE)
mech_diag_dts_bsam <- diagnose_mechanism(
  mech_dts_bsam, dts_treatment, bsam_wet,
  label_x = 'mechanism.dts.bsam.wet.dts')

mechanism_diag <- bind_rows(
  mech_diag_1,
  mech_diag_bsam_wet, mech_diag_bsam_ew,
  mech_diag_dts_ew,   mech_diag_dts_bsam
)
```

```{r combine, include=FALSE}
all_diag <- bind_rows(growth_diag, change_diag, mechanism_diag)

summary_table <- all_diag %>%
  group_by(model) %>%
  group_modify(~ summarize_diag(.x)) %>%
  ungroup() %>%
  arrange(desc(n_both), desc(max_pareto_k))

flagged <- all_diag %>% filter(flag_both)
n_flagged <- nrow(flagged)
n_total   <- nrow(all_diag)
```

# Outlier rule

We flag a measurement as a candidate outlier when *both* of the following hold:

- pointwise WAIC penalty $> 1$, and
- PSIS Pareto $k > 0.5$.

Both thresholds are conservative — meeting either alone is not unusual, and
requiring both jointly identifies observations that the model handles
appreciably worse than its average observation. A flag is not by itself
grounds for exclusion. Flagged observations are excluded only when they are
few enough that exclusion does not amount to reshaping the data to suit the
model, and when the model can be said to fail to fit them rather than simply
to predict them less well than average.

This document applies the rule to all 27 Bayesian models fit in the project — 3
growth models (OASIS, WSAS, engagement), 14 univariate change models for
candidate mechanisms, and 10 multivariate mechanism models linking OASIS to
individual processes. For each model we computed pointwise WAIC and PSIS using
the posterior log-likelihood stored at fit time, flagged observations exceeding
each threshold and those exceeding both, and inspected whether flagged
observations clustered on particular participants, waves, items, or conditions.

# Coverage note for mechanism models

Multivariate mechanism models contain two likelihood statements (OASIS on the
x-side and the candidate process on the m-side). Pointwise log-likelihoods for
these models are available only for the x-outcome, because `ulam`'s `log_lik =
TRUE` option records log-likelihood values only for the first declared likelihood
statement. Per-observation fit for the process-side outcome of each mechanism
model is covered instead by the corresponding univariate change model in the
preceding section, which uses an identical measurement model on the same
observations. Every observation in every model is therefore covered by the
diagnostic through at least one fit.

# Results

## Overall picture

Across all 27 models and `r format(n_total, big.mark=",")` observations,
`r n_flagged` observations (`r round(100*n_flagged/n_total, 3)`%) met the
joint flag (penalty $> 1$ AND $k > 0.5$). No model exceeded a small
handful of flagged observations, and flags were dispersed across measures,
participants, items, and conditions rather than concentrated in any
structural location. The largest concentration in any single model was in
the change model for the TAF-Likelihood subscale, addressed in its own
section below.

## Per-model summary

Table 1 gives the full per-model summary, sorted by joint-flag count.
Most models produced zero to a handful of jointly flagged observations;
the maximum rate was 0.95% (TAF-Likelihood).

\begin{landscape}

```{r summary-table}
summary_table %>%
  kable(
    digits = 3,
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    caption = "Per-model diagnostic summary. n\\_both = observations meeting the joint flag (WAIC penalty > 1 AND Pareto k > 0.5). Models are sorted by n\\_both (descending)."
  ) %>%
  kable_styling(latex_options = c("repeat_header"),
                font_size = 9)
```

\end{landscape}

## Observations meeting both criteria

Table 2 lists the `r n_flagged` observations that met both flag thresholds
(WAIC penalty $> 1$ AND Pareto $k > 0.5$).

```{r flagged-both}
flagged %>%
  select(model, id, condition, wave, question, value, p_waic, pareto_k) %>%
  arrange(desc(pareto_k), desc(p_waic)) %>%
  kable(digits = 3, format = "latex", booktabs = TRUE, longtable = TRUE,
        caption = "Observations exceeding both thresholds (WAIC penalty > 1 AND Pareto k > 0.5).") %>%
  kable_styling(latex_options = c("repeat_header"), font_size = 9)
```

## Clustering

Flagging was dispersed. No single participant accounted for more than three
flagged observations across all 27 models, no single item dominated, and waves
1 and 4 (where the full measurement battery was administered, contributing the
most observations) carried the bulk of flags as expected. Mechanism-model
flags at wave 6 reflect a single follow-up OASIS response by participant 100
that was also flagged in the primary growth model and surfaces in multiple
mechanism fits because each mechanism model includes the full OASIS observation
set.

```{r clustering-id, eval = n_flagged > 0}
flagged %>% count(id, sort = TRUE) %>% head(10) %>%
  kable(caption = "Top 10 participants by number of flagged observations.",
        booktabs = TRUE)
```

```{r clustering-wave, eval = n_flagged > 0}
flagged %>% count(wave, sort = TRUE) %>%
  kable(caption = "Flagged observations by wave.", booktabs = TRUE)
```

```{r clustering-cond, eval = n_flagged > 0}
flagged %>% count(condition, sort = TRUE) %>%
  kable(caption = "Flagged observations by condition. \"NA\" rows correspond to OASIS observations that appear in multiple mechanism-model fits.",
        booktabs = TRUE)
```

```{r clustering-fam, eval = n_flagged > 0}
flagged %>% mutate(family = str_extract(model, '^[^.]+')) %>%
  count(family, sort = TRUE) %>%
  kable(caption = "Flagged observations by model family.", booktabs = TRUE)
```

## Distributions

Figures 1 and 2 show the full distributions of Pareto $k$ and WAIC pointwise
penalty across all models. Both distributions are strongly concentrated near
zero with long, thin right tails.

```{r k-dist, fig.width=8, fig.height=4, fig.cap="Distribution of PSIS Pareto k across all models. Dashed line marks 0.5 (the preregistered flag threshold)."}
ggplot(all_diag, aes(x = pareto_k)) +
  geom_histogram(binwidth = 0.05, boundary = 0, fill = 'grey70', color = 'black') +
  geom_vline(xintercept = 0.5, linetype = 'dashed', color = 'darkorange') +
  labs(x = 'Pareto k', y = 'Count') +
  theme_minimal()
```

```{r w-dist, fig.width=8, fig.height=4, fig.cap="Distribution of pointwise WAIC penalty across all models. Dashed line marks 1.0 (joint flag)."}
ggplot(all_diag, aes(x = p_waic)) +
  geom_histogram(binwidth = 0.1, boundary = 0, fill = 'grey70', color = 'black') +
  geom_vline(xintercept = 1.0, linetype = 'dashed', color = 'darkorange') +
  labs(x = 'WAIC penalty', y = 'Count') +
  theme_minimal()
```

# Decision

The diagnostics support retaining all observations in all models. Across 27
models and `r format(n_total, big.mark=",")` observations, only `r n_flagged`
(`r round(100*n_flagged/n_total, 3)`%) met the conservative joint flag
preregistered for this study. No single model exceeded 1% of flagged
observations. Flags were dispersed across measures, participants, items, and
conditions; no model showed the kind of structural misfit that would call
the model itself into question.

The "if few" qualifier therefore applies. Excluding the flagged observations
would not materially change any reported estimate, and excluding them
selectively risks reshaping the data to suit the model. We retain all
observations in all analyses.

# Note on the TAF-Likelihood measure

One measure warrants comment. The change model for the Thought-Action Fusion
Likelihood subscale (TAF-L) produced noticeably more flagged observations than
any other model in the set — 11 observations met both criteria, 0.95% of that
model's observations, the highest rate in the set. Inspecting Table 2 shows
that these flagged responses are typically extreme low-end responses (value = 1)
on moderately discriminating items.

The TAF-L measure asks participants to endorse the belief that thinking certain
thoughts increases the likelihood of them occurring — a construct with a
strongly skewed distribution in non-clinical samples, where most participants
legitimately endorse near-floor responses on most items. The measurement model
uses a graded response IRT specification that assumes a reasonably continuous
latent trait distribution; the floor effect in TAF-L responses makes individual
extreme-endorsement patterns harder to predict from the overall participant
trait estimate than comparable patterns on less-skewed measures. This is a
known limitation of IRT models applied to skewed constructs rather than a
problem specific to our fit, and the model remains usable.

TAF-L results should accordingly be interpreted with slightly more caution than
results from the other measures. We note this here for completeness; TAF-L was
not identified as a mechanism of change in the present analyses.

# Artifacts

Per-observation diagnostics for all models, the per-model summary table, and
the table of flagged observations have been saved as CSV and RDS files under
`data/diagnostics/` in the repository for independent verification.

```{r save, include=FALSE}
dir.create(here('data', 'diagnostics'), showWarnings = FALSE, recursive = TRUE)
saveRDS(all_diag,      here('data', 'diagnostics', 'outlier_diagnostics_all.rds'))
saveRDS(summary_table, here('data', 'diagnostics', 'outlier_summary.rds'))
write_csv(summary_table, here('data', 'diagnostics', 'outlier_summary.csv'))
write_csv(flagged,       here('data', 'diagnostics', 'outliers_flagged.csv'))
```
