# Transparency and Openness

## Preregistration:
This study was preregistered prior to data collection at [doi.org/10.17605/OSF.IO/2CEHR](doi.org/10.17605/OSF.IO/2CEHR).

## Deviations from preregistration:
We departed from the preregistration on the following points; all decisions were made before data collection.

- We added a passive waitlist control group, drawn from a prior study with identical inclusion, exclusion, and procedural criteria on the same recruitment platform [see @sorkaTransdiagnosticSelfGuidedImageryinpreparation]. Because this group was not randomized concurrently with the intervention arms, results involving it should be interpreted as comparative rather than inferential.
- We administered follow-up assessments at one week and two months rather than at a single unspecified follow-up.
- We replaced the Thought Fusion Inventory with the Thought-Action Fusion Scale Revised, administered pre- and post-intervention rather than at every session.
- State anxiety was measured using the Brief State Anxiety Measure (instrument unspecified in the preregistration).
- We added the Core Threat Structured Interview (intervention component) and the Treatment Credibility Questionnaire (manipulation check).
- We specified the growth model's time component as piecewise rather than as a single linear slope, reflecting the qualitatively different change regimes across the intervention and the two follow-up windows.
- The Imaginal Behavior Approach Test was administered as preregistered but not analyzed in the present report.

## Data, materials, code, and online resources:
Data, analysis scripts, and methodological documentation are available at [https://github.com/eladzlot/transdiagnostic.wet-2025-public](https://github.com/eladzlot/transdiagnostic.wet-2025-public).
Datasets include only quantitative information to protect participant confidentiality.

## Reporting:
We report how we determined our sample size, all data exclusions, all manipulations, and all measures in the study.
A power analysis indicated a target of 148 participants.
Recruitment was halted before this target was reached under a preregistered Bayesian stopping rule; specifics are given in the Methods.

## Ethical approval:
This research was approved by the Ethics Committee of The Hebrew University of Jerusalem.
All participants provided informed consent.

# Methods

This study was a randomized controlled trial (RCT) in which participants were randomly assigned to one of two groups: the treatment group (WET) or the control group (Neutral Imagery Writing, NIW).
Additionally, data from a waitlist control group was incorporated, collected from a prior study with the same structure and inclusion/exclusion criteria [see @sorkaTransdiagnosticSelfGuidedImageryinpreparation].
This provided both a passive control and an active control for comparison.
The study employed a double-blind design to ensure objectivity.
Study personnel had no direct interaction with participants, and both experimenters and participants were unaware of group assignments until the one-week follow-up.
The primary outcome was defined as the change in OASIS scores, with changes in functioning scores and other proposed mechanisms serving as secondary outcomes.
Covariations between changes in OASIS and changes in each potential mechanism were also examined.

## Participants

Participants were recruited through the Prolific platform, compensated at $\pounds 9$/hour.
Inclusion criteria included moderate pathological anxiety (OASIS score > 4)
and impaired daily functioning (at least one WSAS item scored above 2 on a 0–8 scale).
Exclusion criteria were severe depression [greater than 14 on the PHQ\; @kroenkePHQ8MeasureCurrent2009], elevated post-traumatic symptoms [greater than 6 on the brief PCL-5\; @zuromskiDevelopingOptimalShortform2019], or presence of psychotic symptoms [evaluated via items 19 and 20 from the DIAMOND screener\; @tolinPsychometricPropertiesStructured2018].
Furthermore, participants who had reported head injury or reading and writing difficulties in the Prolific system were excluded.
In addition, English as a fluent language was required.
Finally, only experienced users on Prolific with an approval rate above 95, and minimum 300 previous submissions were allowed to participate.
An ethnically diverse sample was requested from Prolific.

Power analysis indicated that a sample size of 148 participants would provide 0.8 power.
Due to financial constraints, we preregistered a Bayesian stopping rule allowing for early termination: participants were recruited in increments of 30, and after each increment we evaluated whether 90% of the posterior distribution for the WET–NIW difference in end-of-treatment OASIS fell within a region of practical equivalence of ±2 points.
Recruitment stopped when this criterion was met.

Participants were randomly assigned to the treatment and control groups in a 2:1 ratio, using stratified randomization ($OASIS < or \geq 8$), aiming to ensure equal distribution of anxiety severity between groups.
The block randomization process was integrated into the experiment program code.

## Measures

```{r reliability, include=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
library(psych)
source(here('scripts','helpers.R'))
if (!exists('data_long')) source(here('scripts','load_data.R'))

measure_list = c('oasis', 'wsas', 'dts','taf_m','taf_l', 'bcss_self_neg', 'bcss_other_neg','bcss_self_pos', 'bcss_other_pos','tiii','ifs', 'scs', 'mcq_confidence','mcq_worry','mcq_self','mcq_danger','mcq_control', 'bsam.pre', 'bsam.post')

omega.list = list()
for (m in measure_list) {
  omega.list[[m]] = omega_from_long(data_long %>% filter(measure == m))
}
omega.list[['bsam']] = omega_from_long(data_long %>% filter(measure %in% c('bsam.pre', 'bsam.post')))
```

The *Overall Anxiety Severity and Impairment Scale* [OASIS\; @normanDevelopmentValidationOverall2006] assesses the frequency, intensity, and functional impairment of anxiety and fear over the past week.
Participants rate their anxiety on a 5-point scale from 0 (little or none) to 4 (extreme or all the time).
Higher scores indicate more severe anxiety-related impairment, with a cut-off score of 8 recommended for identifying anxiety disorders and a change of four points considered clinically significant [@moorePsychometricEvaluationOverall2015].
The OASIS has strong psychometric properties, with an omega coefficient of `r omega.list$oasis` across timepoints in our sample, indicating high reliability.

The *Work and Social Adjustment Scale* [WSAS\; @mundtWorkSocialAdjustment2002] was used to assess functional impairment.
The scale consists of five items, each rated on a 9-point Likert scale (0 = no impairment to 8 = severe impairment), evaluating the extent of difficulty in work, home management, social leisure, private leisure, and close relationships.
In our sample, the omega coefficient  was `r omega.list$wsas`, indicating high reliability.

The *Thought-Action Fusion Scale Revised* [TAF-R\; @shafranThoughtactionFusionObsessive1996] measured the degree to which individuals equate thoughts with actions.
It includes 19 items rated on a 5-point Likert scale (0 = disagree strongly to 4 = agree strongly), with higher scores indicating a stronger tendency toward thought–action fusion.
The scale comprises two subscales: TAF-Moral, which reflects the belief that having unacceptable thoughts is morally equivalent to performing them, and TAF-Likelihood, which captures the belief that thinking certain thoughts increases the likelihood of them occurring.
We analyzed these subscales separately.
In our sample, the omega coefficients for moral and likelihood TAF were `r omega.list$taf_m` and `r omega.list$taf_l` respectively, indicating excellent reliability.

The *Brief Core Schema Scale* [BCSS\; @fowlerBriefCoreSchema2006] was employed to assess cognitive schemas.
It consists of 24 items regarding the self or others (e.g., “I am talented”; “I am loved”), divided into two subscales each (e.g., self-positive and self-negative).
Each item is rated on a 5-point Likert scale (0 = Don’t Believe it to 4 = Believe it totally).
In our sample, the omega coefficients for each scale were at least `r min(c(omega.list$bcss_self_pos, omega.list$bcss_self_neg, omega.list$bcss_other_pos, omega.list$bcss_other_neg))`, indicating excellent reliability.

The *Distress Tolerance Scale* [DTS\; @simonsDistressToleranceScale2005] assessed participants' ability to tolerate distressing emotions.
The DTS consists of 15 items rated on a 5-point Likert scale (1 = strongly agree to 5 = strongly disagree), evaluating various aspects of distress tolerance.
Due to a coding mistake, item 6 was missing and item 7 was repeated twice, thus we dropped item 6 completely from all analyses.
In our sample, the omega coefficient was `r omega.list$dts`, indicating excellent reliability.

The *Metacognition Questionnaire-30* [MCQ-30\; @wellsShortFormMetacognitions2004] was employed to evaluate metacognitive beliefs related to anxiety.
The MCQ-30 consists of 30 items rated on a Likert scale ranging from 1 (do not agree) to 4 (agree very much).
It assesses five factors of metacognition: cognitive confidence, positive beliefs about worry, cognitive self-consciousness, negative beliefs about the uncontrollability of thoughts and danger, and beliefs about the need to control thoughts.
In our sample, the omega coefficients  were at least `r min(c(omega.list$mcq_danger, omega.list$mcq_confidence, omega.list$mcq_worry, omega.list$mcq_self, omega.list$control))`, indicating excellent reliability.

The *Self-Compassion Scale-Short Form* [SCS-SF\; @raesConstructionFactorialValidation2011] assessed participants’ levels of self-compassion.
The SCS-SF consists of 12 items rated on a 5-point Likert scale (1 = almost never to 5 = almost always).
Higher scores indicate a greater frequency of self-compassionate thoughts and behaviors.
In our sample, the omega coefficient  was `r omega.list$scs`, indicating excellent reliability.

The *Impact of Future Event Scale* [IFES\; @deeproseMeasuringIntrusiveProspective2011] assessed the impact of intrusive prospective, personally-relevant imagery.
Participants completed 24 items evaluating intrusive pre-experiencing, avoidance, and hyper-arousal, responding to how frequently each statement was true for them over the past seven days.
Responses were given on a 5-point scale (0 = not at all to 4 = frequently).
In our sample, the omega coefficient  was `r omega.list$ifs`, indicating excellent reliability.

The *Brief State Anxiety Measure*  [BSAM\; @bergAreEmotionsFrightening1998] was used to assess participants' state anxiety in response to their core threat. 
The BSAM consists of six items (relaxed, steady, strained, comfortable, worried, tense) adapted from the State-Trait Anxiety Inventory.
These items are designed to provide a brief and efficient measure of state anxiety.
Each item is rated on a 7-point scale (1 = Not at all, 7 = Very much so), with higher scores indicating greater levels of anxiety.
In this study, participants were asked to imagine their core threat script for a few moments both before and after each writing session.
They then completed the BSAM to report their current anxiety levels.
The BSAM demonstrated excellent internal consistency in our sample, as indicated by an omega coefficient of `r omega.list$bsam`.

The *Treatment Credibility Questionnaire* [@devillyPsychometricPropertiesCredibility2000] was used to evaluate treatment credibility.
Credibility was assessed once, at the middle of Session 1, following the presentation of the overall treatment rationale and before the intervention itself.
Patients rated the therapy on a scale from 0 (not at all) to 8 (very), assessing both how they felt about the intervention and how effective they believed it would be.
The scale has demonstrated strong internal consistency and test–retest reliability [@devillyPsychometricPropertiesCredibility2000].

```{r engagement.methods, echo=FALSE, include=FALSE, cache=TRUE}
library(psych)
raw.engagement = read_csv(here('raw/engagement.csv'))

kap = raw.engagement %>% select(-id) %>%
  filter(complete.cases(.)) %>%
  mutate(across(everything(), as.logical)) %>%
  as.matrix %>%
  cohen.kappa()


engagement_kappa = sprintf(
  "Cohen’s κ = %.2f, 95%% CI [%.2f, %.2f], %.1f%% agreement",
  kap$kappa,
  kap$confid[1,1], kap$confid[1,3],
  sum(diag(kap$agree)) * 100
)
```

Following reviewer feedback, we conducted a manualized content-coding procedure to assess core-threat engagement in the writing exercises.
Two independent judges classified each participant’s four-wave exercise set as either *engaged* or *not engaged*, using a conservative binary rule: participants were coded as engaged if at least one wave explicitly and emotionally addressed their pre-identified core threat (see supplementary materials).
All participants were coded by the first judge, and a random subset of 20 participants (40%) was independently coded by the second judge to evaluate inter-rater reliability.
Inter-rater agreement was substantial (`r engagement_kappa`).

## Procedure

After screening, participants completed four 20-minute writing sessions spaced approximately two days apart.
Follow-up assessments were conducted one week and two months after completion of the final session.
The waitlist control group completed questionnaires at time points equivalent to the active conditions’ initial session, fourth session, and one-week follow-up.

Anxiety symptoms, measured using the OASIS, were assessed before each writing session, except for the final session, in which symptoms were assessed immediately after writing.
Additionally, symptoms were measured at both follow-up sessions.
Functional impairment, measured using the Work and Social Adjustment Scale (WSAS), was assessed at pre-intervention, post-intervention, and at both follow-ups.
Mechanistic variables were assessed at pre-intervention (prior to the first writing session) and post-intervention (after the fourth writing session).

Before the first writing session, participants identified their primary core threat using the Core Threat Structured Interview (CTSI).
The CTSI is a structured and psychometrically validated online assessment designed to elicit the individual’s ultimate feared outcome underlying anxiety or avoidance [@zlotnickCoreThreatStructured2026].
It guides participants through examples and clarification prompts to ensure that the selected threat is personally meaningful and specific enough to serve as the focus for the writing exercises.
Finally, participants developed a concise script creating a clear narrative of how these threats were expected to unfold from their proximal threats.

Participants were randomized into two groups: WET and NIW, they were later compared with a waitlist group.

## Intervention

**Intervention Development and Acceptability:**
The intervention was developed iteratively through repeated pilot administrations (two to three participants per cycle), each followed by structured feedback regarding clarity, difficulty, and engagement.
Revisions were made across cycles until the protocol was judged acceptable for implementation.
Acceptability and perceived usefulness were assessed in the present study using post-intervention and follow-up self-report items.

**Written Exposure Therapy (WET).**
Participants received brief exposure-based psychoeducation emphasizing that repeated engagement with distressing material reduces emotional impact [cf. @benitoMechanismsChangeExposure2024].
They then repeatedly expanded their personalized threat script through detailed writing focused on feared content, attending to their thoughts and emotions.

**Neutral Imagery Writing (NIW).**
Participants received psychoeducation emphasizing distraction and mental control in emotional regulation and were guided to describe a neutral situation using vivid sensory detail while avoiding emotional or threat-related content.

**Waitlist Control.**
The waitlist group served as an inactive control and completed the OASIS and WSAS at three one-week intervals corresponding to the pre-intervention, post-intervention, and one-week follow-up assessments of the active conditions.

Full instructions for the WET and NIW conditions, including psychoeducation materials, are provided in the supplementary materials.

## Statistical Analysis

**Change Model:** We employed a multilevel ANCOVA design to account for the nested structure of the data, where anxiety is measured repeatedly within individuals [@bodnerDetectingDifferentiatingDirection2018; @vanbreukelenANCOVACHANGEBaseline2013].
By accounting for both within- and between-subject variability, multilevel modeling reduces bias providing more precise estimates of treatment effects.
Furthermore, it becomes possible to investigate individual as well as group level effects [@raudenbushHierarchicalLinearModels2002].

To determine treatment efficacy we examined the *fixed effects* — the average rates of change across individuals.
As described below, to understand covariation between symptoms and processes we examined the *random effects* — the individual deviations from the group-level trends.
To account for follow-up measurements at varying time lags (e.g., 1 week, 2 months), we employ a piecewise growth model, allowing for separate slopes at different follow-up times [@bollenLatentCurveModels2006; @andersson35yearFollowInternetdeliveredCognitive2013].
Combined with the measurement model described below, this is in essence a latent change score model [@cancerDynamicalPropertiesConceptual2021; @mcardleLatentVariableModeling2009].

The equation used was:

$$\theta_{ti} = \beta_{0i} + \beta_{1i} \cdot \min(t-1, 3) \cdot 1_{\{t > 1\}} + \beta_{2i} \cdot 1_{\{t > 5\}} + \beta_{3i} \cdot 1_{\{t=6\}}$$
$$\beta_{ji} = \bar{\beta}_{j,\text{condition}} + u_{ji}$$

The latent symptom severity ($\theta$) was linearly modeled as a baseline effect ($\beta_{0i}$) for each individual and the rate of change between the first and fourth sessions ($\beta_{1i}$).
Separate terms ($\beta_{2i}$) and ($\beta_{3i}$) represent the rate of change at follow-up (1 week, 2 months).
The individual-specific effects, ($\beta_{ji}$), are modeled as deviations from the group-level effects, ($\bar{\beta}_{j,\text{condition}}$).

**Measurement Model:** Given that anxiety was measured using an ordinal scale, we linked latent symptom severity ($\theta$) to the observed ordinal responses using an ordered logit model with a cumulative link function [@mccullaghRegressionModelsOrdinal1980].
This model assumes that ordinal responses are generated by an underlying continuous latent variable, with thresholds ($\kappa$) defining the response categories.
The ordered logit function estimates the probability ($\phi$) of endorsing each response category based on the latent symptom severity trait ($\theta$).

To account for measurement error and the multidimensionality of the anxiety construct, we adopted a graded response model (GRM) from item response theory (IRT) [@samejimaGradedResponseModels2016].
The GRM is well-suited for ordinal data, modeling the probability of endorsing an item category as a function of both item-specific parameters (discrimination: ($\alpha$), and difficulty: ($\beta$) and the individual’s latent trait level ($\theta$).

The GRM allows us to model the latent symptom severity ($\theta$) for each individual at a specific time point, while accounting for the unique characteristics of each item.
The discrimination parameter ($\alpha$) captures how well each item distinguishes between individuals with different levels of anxiety, while the difficulty parameter ($\beta$) reflects the trait level required to endorse higher response categories.
This approach helps address known issues with the reliability of change scores [@prielerProblemsMeasurementChange2008; @rodebaughUnreliabilityThreatUnderstanding2016].

In this framework, $\theta$ is estimated from the multilevel model described earlier, capturing both individual-specific (random) effects and group-level (fixed) effects.
The integration of the GRM with the broader multilevel model allows us to model the latent trait accurately, taking into account both item-level variability and individual differences in anxiety over time.

The equation for the measurement model is as follows:

$$R_{itq} \sim \text{Ordered-logit}(\phi_{itq}, \kappa)$$
$$\phi_{itq} = \alpha_q \cdot (\theta_{it} - \beta_q)$$

Here, the likelihood function represents the probability of observing the ordinal response ($R$), given the latent symptom severity  ($\theta$).
The IRT model specifies the relationship between the latent trait and the item parameters.
Detailed specifications for this model and the full code is available in the supplementary materials.

**Mechanisms:** Our second research question investigates the role of mechanisms in change in pathological anxiety.
Following Kazdin’s [-@kazdinMediatorsMechanismsChange2007] framework for mechanisms research, the present study intentionally employed a broad exploratory approach to identify potential targets for future confirmatory work.
While establishing causality is difficult [see @eronenCausalDiscoveryProblem2020], our goal is to identify correlations that may signal candidate causal relationships.

Mechanisms of change in psychotherapy are processes or events through which psychological interventions affect symptoms or functioning.
Demonstrating that an intervention causes change does not explain *why* this change came about.
Establishing that a process is a mechanism requires association between the intervention and process, and between the process and the symptoms [@kazdinMediatorsMechanismsChange2007].
Thus, we first establish which processes are affected by the intervention.
Next, we examine the relationship between the process and the outcome.

To identify processes that may mediate change, we examined both fear-specific and fear-agnostic mechanisms previously linked to exposure-based interventions.
Reductions in fear, indexed by state anxiety during imagining the core threat [BSAM\; @bergAreEmotionsFrightening1998], reflect habituation or extinction processes.
Broader regulatory and cognitive mechanisms included distress tolerance [DTS\; @simonsDistressToleranceScale2005], reflecting the ability to endure negative emotion, self-compassion [SCS\; @raesConstructionFactorialValidation2011], representing adaptive self-relating and emotional regulation; and thought–action fusion [TAF-R\; @shafranThoughtactionFusionObsessive1996], the belief that thoughts can cause or prevent real outcomes.
Metacognitive beliefs [MCQ-30\; @wellsShortFormMetacognitions2004] and core schemas [BCSS\; @fowlerBriefCoreSchema2006] indexed higher-order beliefs about control, self, and others.
Finally, imagery-specific reactions [IFES\; @deeproseMeasuringIntrusiveProspective2011] captured distress, intrusiveness, and avoidance in response to personally salient imagery.
Together, these constructs span both targeted and transdiagnostic mechanisms, allowing a comprehensive view of how WET may promote change across emotional, cognitive, and regulatory domains.

We focus on the process of change, thus we're interested in the association between *changes* in processes and *changes* in symptoms.
To examine this association we use multivariate latent curve models [@maccallumStudyingMultivariateChange1997; @damianDoesVariabilityHuman2010; @baldwinAnalyzingMultipleOutcomes2014].
This approach extends  univariate growth models by allowing the simultaneous examination of multiple outcomes, capturing both the relationships between individual processes and how they evolve over time.
Specifically, we model both anxiety (as measured by the OASIS) and the process of interest in parallel, using a single covariance matrix to estimate the random effects for both processes.
By incorporating a shared covariance matrix, we capture the correlations between baseline values (intercepts) and changes (slopes) both within and across the two outcomes.
This enables us to estimate the correlation between changes controlling for baseline values and accounting for measurement error.
Detailed specifications for this model and the full code are available in the supplementary materials.

Although this analysis does not establish causality, it offers valuable insights into potential intervention targets.
For example, identifying processes that strongly correlate with changes in anxiety could inform more focused treatments by highlighting which processes to target for clinical improvement.
Further research will need to extend these findings by extending them into full mediation models [@hilleyDynamicChangeMeets2022].

### Missing Data and Estimation

All participants who began the intervention were included in the analysis.
Missing data were imputed via Bayesian imputation, assuming that missingness may be affected by baseline trait levels and task completion difficulty.

We used Bayesian methods to estimate the parameters of the model using Markov Chain Monte Carlo (MCMC) algorithms implemented in Stan [@carpenterStanProbabilisticProgramming2017], and via the rethinking interface [@mcelreathStatisticalRethinkingBayesian2020].
We specified weakly informative priors for the regression coefficients, the thresholds, and the item parameters.
We assessed the convergence and the fit of the model using posterior predictive checks, trace plots, and effective sample sizes.
We additionally assessed measurement-level fit using pointwise WAIC and PSIS: observations with WAIC pointwise penalty greater than 1 *and* PSIS Pareto $k$ greater than .5 were flagged for inspection, with potential exclusion if such observations were few and the model failed to fit them.
Across all 27 models in the study, flagged observations were rare (under 1% of all measurements; under 1% in any single model) and dispersed across measures, participants, and items; we therefore retained all observations.
Per-model diagnostics are reported in the supplementary material.
We use a credible interval of 89% [@mcelreathStatisticalRethinkingBayesian2020].
