```{r results.setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(rethinking)
library(here)
library(papaja)
source(here('scripts','load_data.R'))
source(here('scripts','model.helpers.R'))
```

# Results

## Participants

```{r participants, echo=FALSE, include=FALSE}

N.screened = screening$prolific_pid %>% unique %>% length()
N.randomized  = pid.condition %>% filter(condition != 'waitlist') %>% nrow()
N.treatment  = pid.condition %>% filter(condition == 'treatment') %>% nrow()
N.control  = pid.condition %>% filter(condition == 'control') %>% nrow()
N.waitlist = pid.condition %>% filter(condition == 'waitlist') %>% nrow()


d = demographics.data %>% filter(condition != 'waitlist')
# checking condition differences
conditions.demog.ps = max(c(
    fisher.test(table(d$condition, d$Ethnicity))$p.value,
    fisher.test(table(d$condition, d$Sex))$p.value,
    wilcox.test(oasis ~ condition, data = d)$p.value,
    wilcox.test(wsas ~ condition, data = d)$p.value
))
```

Recruitment began August 6, 2023, and ended September 25, 2023, and the final assessment occurred December 2, 2023.
Of `r N.screened` individuals initially screened,
`r N.randomized` were randomly assigned at a 2:1 ratio to WET `r format_count_perc(N.treatment, N.randomized)` or NIW `r format_count_perc(N.control, N.randomized)` conditions.
The age of participants in the NIW group was significantly lower than that of participants in the WET group (Wilcoxon test, `r apa_print(wilcox.test(Age ~ condition, data = d))$full_result`). 
Fisher exact test indicated no significant group differences in ethnicity or sex, and Wilcoxon tests showed no significant group differences for pre-treatment symptom measures ($ps >$ `r conditions.demog.ps`).

Given the similarity in post-treatment outcomes between the treatment and control groups, we incorporated additional data (n = `r N.waitlist`) from a waitlist control group as an idiographic benchmarking group.
This data was collected in a similar study conducted between February 1 and February 24, 2023 [@sorkaTransdiagnosticSelfGuidedImageryinpreparation].
This was done in order to make sure that the changes found during the interventions were not due to chance.
These participants were randomly assigned to a waitlist control group for a study that focused on mechanisms of imagery rescripting.
Given the identical inclusion/exclusion criteria and the minor time gap between data collection periods, we expect no systematic differences between the groups.
The CONSORT flowchart (see Figure \@ref(fig:consort)) provides a detailed overview of the recruitment, randomization, and follow-up processes across the two active groups.
Comprehensive demographic information is provided in Table \@ref(tab:demographics).

(ref:consort) A CONSORT flowchart providing a detailed overview of the recruitment, randomization, and follow-up processes across the two active groups.

```{r consort, echo=FALSE, fig.cap='(ref:consort)', fig.height=6, fig.width=6, message=FALSE, warning=FALSE, cache=TRUE, file=here('scripts','figures.consort.R'), out.width="100%"}
```


```{r completed, echo=FALSE, include=FALSE}
d <- demographics.data %>% filter(condition !='waitlist')

# Calculate completion statistics
N.completed <- demographics.data$completed %>% sum
N.completed.treatment <- demographics.data %>% filter(condition == "treatment", completed==TRUE) %>% nrow()
N.completed.control <- demographics.data %>% filter(condition == "control", completed==TRUE) %>% nrow()

# Perform statistical tests
completed.test <- table(d$condition, d$completed) %>% fisher.test()

# Checking demographic differences for completers
completed.sex.table <- table(d$completed, d$Sex)
completed.demog.ps <- min(c(
  fisher.test(table(d$completed, d$Ethnicity))$p.value,
  wilcox.test(Age ~ completed, data = d)$p.value,
  wilcox.test(oasis ~ completed, data = d)$p.value,
  wilcox.test(wsas ~ completed, data = d)$p.value
))
```

The large majority of participants `r format_count_perc(N.completed, N.randomized)` completed all assessments.
Study completion rates did not differ between treatment (`r format_perc(N.completed.treatment, N.treatment)`) and control (`r format_perc(N.completed.control, N.control)`); `r apa_print.fisher(completed.test)`.
Female participants had a significantly lower attrition rate (`r format_perc(completed.sex.table[1, 1], sum(completed.sex.table[, 1]))`) compared to males (`r format_perc(completed.sex.table[1, 2], sum(completed.sex.table[, 2]))`); `r apa_print.fisher(fisher.test(completed.sex.table))`.
Fisher’s exact and Wilcoxon tests revealed no significant differences between study completers and non-completers in terms of age, ethnicity, or symptom measures ($ps > `r completed.demog.ps`$).

(ref:demographics) Pre-treatment and Demographic Characteristics of Intent-to-Treat Sample. WET = Written Exposure, NIW = Neutral Imagery Writing, Waitlist group were added as a reference from a different study [@sorkaTransdiagnosticSelfGuidedImageryinpreparation].

```{r demographics, file=here('scripts','tables.demographics.R'), echo=FALSE, message=FALSE, warning=FALSE,cache=TRUE}
```

## Intervention Quality
```{r credibility, echo=FALSE, include=FALSE, cache=TRUE}
# Filter and summarize the data for credibility measures
d <- data_long %>%
  filter(measure == 'credibility.think' | measure == 'credibility.feel') %>%
  group_by(measure, condition, id) %>%
  summarise(value = sum(value), .groups = "drop")

# Extract cognitive scores for treatment and control conditions
cognitive_score_wet <- d %>%
  filter(condition == 'treatment', measure == 'credibility.think') %>%
  pull(value)

cognitive_score_ew <- d %>%
  filter(condition == 'control', measure == 'credibility.think') %>%
  pull(value)

# Extract feeling scores for treatment and control conditions
feeling_score_wet <- d %>%
  filter(condition == 'treatment', measure == 'credibility.feel') %>%
  pull(value)

feeling_score_ew <- d %>%
  filter(condition == 'control', measure == 'credibility.feel') %>%
  pull(value)
```

Both WET and NIW were rated highly on the Devilly Credibility Scale, indicating strong credibility in both interventions.
On the cognitive subscale, WET scored `r format_mean_sd(cognitive_score_wet)`, while NIW scored  `r format_mean_sd(cognitive_score_ew)`.
Similarly, on the feeling subscale, WET scored  `r format_mean_sd(feeling_score_wet)` and NIW scored `r format_mean_sd(feeling_score_ew)`.
These results suggest that participants found both interventions to be highly credible in terms of both their perceived logic and their intuitive appeal.
In their open feedback after the study, participants generally found the writing task acceptable, though some in the WET condition reported it as difficult, while some in the NIW condition described it as progressively boring.

Participants in the WET condition successfully created powerful exposure scripts that addressed their core threats.
For instance, an excerpt from the script of a socially anxious individual reads:

> What am I doing here? I am a nobody. I know nothing. I feel almost sick now with nerves and dread. What if I just left now? Would anyone even notice? But it's too late…  
> I am to speak first, so I can't even steal other people's ideas…  
> [They're thinking] Who is this idiot? I stumble on and on. I want to stop and for it all to be over, but I can't. It just goes on forever. It is like dying in slow motion.

Adherence in the NIW condition was less confined, with some participants writing scripts that focused on positive outcomes rather than neutral events.
For example, one participant wrote:

> I feel enthusiastic about starting my day on a positive and productive day.
> Today instead of going through my phone for an hour on social media, I start with making my bed and cleaning my room.
> ... I start applying for opportunities that align with my qualifications and a bit of me is worried but still hopeful that I get recognized and recruited.

This type of writing is similar to what is done in imagery rehearsal [cf. @krakowClinicalManagementChronic2006].

## Symptom Measures

```{r growth.changes, echo=FALSE, include=FALSE, cache=TRUE}
source(here('scripts','model.growth.R'))
d = data_long %>% filter(measure == 'oasis', !is.na(value), wave>0) %>% left_join(demographics.data)
m = model.growth(d, 'oasis')
post = extract.samples(m)

sd.wet = d %>% filter(wave == 1, condition == 'treatment') %>% group_by(id) %>% summarize(v = sum(value)) %>% pull(v) %>% sd
sd.ew = d %>% filter(wave == 1, condition == 'control') %>% group_by(id) %>% summarize(v = sum(value)) %>% pull(v) %>% sd
sd.wl = d %>% filter(wave == 1, condition == 'waitlist') %>% group_by(id) %>% summarize(v = sum(value)) %>% pull(v) %>% sd

expected = list(
  growth_link(post,1),
  growth_link(post,2),
  growth_link(post,3)
)

delta.wet = expected[[1]][,1]-expected[[1]][,5]
delta.ew = expected[[2]][,1]-expected[[2]][,5]
delta.wl = expected[[3]][,1]-expected[[3]][,5]

delta.treatment.wet = (expected[[1]][,1]-expected[[1]][,4])/sd.wet
delta.treatment.ew = (expected[[2]][,1]-expected[[2]][,4])/sd.ew
delta.treatment.wl = (expected[[3]][,1]-expected[[3]][,4])/sd.wl
delta.fu1.wet = (expected[[1]][,4]-expected[[1]][,5])/sd.wet
delta.fu1.ew = (expected[[2]][,4]-expected[[2]][,5])/sd.ew
delta.fu1.wl = (expected[[3]][,4]-expected[[3]][,5])/sd.wl
delta.fu2.wet = (expected[[1]][,5]-expected[[1]][,6])/sd.wet
delta.fu2.ew = (expected[[2]][,5]-expected[[2]][,6])/sd.ew
delta.fu2.wl = (expected[[3]][,5]-expected[[3]][,6])/sd.ew
delta.all.wet = (expected[[1]][,1]-expected[[1]][,6])/sd.wet
delta.all.ew = (expected[[2]][,1]-expected[[2]][,6])/sd.ew
delta.all.wl = (expected[[3]][,1]-expected[[3]][,6])/sd.wl
```

We fit a Bayesian latent growth curve model to assess the efficacy of the interventions (see Figure \@ref(fig:growth) for a visualization of the model's estimated trajectories). 
We calculated the standardized mean change (SMC) for each group using the pre-treatment standard deviation as the denominator [@beckerSynthesizingStandardizedMeanchange1988].
This standardization allows comparisons across measures, with higher effect size values indicating better outcomes.
Additionally, we report the probability of direction, which quantifies the certainty that the outcome is greater than 0.
The probability of direction, marked in this paper as P(>0), ranges from 0 to 1, where values close to 0 indicate high certainty of a negative effect (symptom deterioration), and values close to 1 indicate high certainty of a positive effect [symptom improvement See @makowskiIndicesEffectExistence2019].

(ref:growth) Estimated treatment trajectories for each intervention based on the latent growth model. Values are adjusted for gender. WET = Written Exposure Therapy; NIW = Neutral Imagery Writing. The time intervals between pre-treatment and post-treatment, and between post-treatment and followup 1 (FU1), were approximately one week. The interval between FU1 and followup 2 (FU2) was approximately two months, which is indicated by the dashed regression line. The shaded gray area represents the 89% credible interval (CI).

```{r growth, file=here('scripts','figures.growth.R'), echo=FALSE, message=FALSE, warning=FALSE, fig.cap='(ref:growth)', cache=TRUE}
```

Clinical anxiety, as measured by the OASIS, significantly improved across all three interventions at the one-week follow-up.
The estimated raw score improvements were `r format_estimate(delta.wet, estimand = 'M')` for the WET group, `r format_estimate(delta.ew, estimand = 'M')` for the NIW group, and `r format_estimate(delta.wl, estimand = 'M')` for the Waitlist group.

Participants in the WET group showed significant improvements during the treatment period compared to the Waitlist group (`r format_estimate(delta.treatment.wet - delta.treatment.wl)`).
These gains continued at a similar rate at the one-week follow-up, again compared to the Waitlist group (`r format_estimate(delta.fu1.wet - delta.fu1.wl)`).
At the two-month follow-up, the gains in WET were estimated to continue at a slower rate compared to the previous gains observed in the WET group (`r format_estimate(delta.fu2.wet - delta.fu2.wl)`).
Notably, this rate of change was estimated in relation to the projected Waitlist trajectory, as no Waitlist data was collected at this time point.
The improvement trajectory appeared linear, with consistent gains observed throughout the intervention and continuing during both follow-ups.
Total improvement when compared to the projected Waitlist trend was `r format_estimate(delta.all.wet - delta.all.wl)`.

Participants in the NIW group also demonstrated significant improvements during the treatment period compared to the Waitlist group (`r format_estimate(delta.treatment.ew - delta.treatment.wl)`).
These improvements were sustained at the one-week follow-up, again in comparison to the Waitlist group (`r format_estimate(delta.fu1.ew - delta.fu1.wl)`).
At the two-month follow-up, the gains were maintained relative to the projected Waitlist trajectory, as no Waitlist data was collected at this time point (`r format_estimate(delta.fu2.ew - delta.fu2.wl)`).
Participants in the NIW group showed considerable overall improvement (`r format_estimate(delta.all.ew - delta.all.wl)`).
From pre-treatment to the two-month follow-up, WET demonstrated a slightly greater cumulative advantage over NIW (`r format_estimate(delta.all.wet - delta.all.ew)`).
While participants in the NIW group showed considerable overall improvement (`r format_estimate(delta.all.ew - delta.all.wl)`), the L-shaped trajectory suggests that most of the gains occurred during the treatment phase and then stabilized.
Contrary to our hypothesis, the NIW group had a slight advantage over WET during the treatment phase (`r format_estimate(delta.treatment.wet - delta.treatment.ew)`).
However, WET outperformed NIW at both the one-week follow-up (`r format_estimate(delta.fu1.wet - delta.fu1.ew)`) and the two-month follow-up (`r format_estimate(delta.fu2.wet - delta.fu2.ew)`).

These findings suggest that both WET and NIW interventions effectively reduce trait anxiety compared to the Waitlist control.
However, while NIW showed slightly greater gains during the treatment phase, WET exhibited stronger and more sustained improvements post-treatment.
This pattern highlights the potential of WET for achieving longer-term symptom reduction.

```{r growth.changes.wsas, echo=FALSE, include=FALSE, cache=TRUE}
# Source the growth model script
source(here('scripts', 'model.growth.R'))

# Filter and join the data for WSAS measure
d <- data_long %>%
  filter(measure == 'wsas', !is.na(value), wave > 0) %>%
  left_join(demographics.data, by = c("id", 'condition'))  # Ensure proper join key if needed

sd.wsas.wet = d %>% filter(wave == 1, condition == 'treatment') %>% group_by(id) %>% summarize(v = sum(value)) %>% pull(v) %>% sd
sd.wsas.ew = d %>% filter(wave == 1, condition == 'control') %>% group_by(id) %>% summarize(v = sum(value)) %>% pull(v) %>% sd
sd.wsas.wl = d %>% filter(wave == 1, condition == 'waitlist') %>% group_by(id) %>% summarize(v = sum(value)) %>% pull(v) %>% sd

# Fit the growth model
m <- model.growth(d, 'wsas', force = FALSE)

# Extract posterior samples
post <- extract.samples(m)

# Compute expected growth links
expected <- list(
  growth_link(post, 1),
  growth_link(post, 2),
  growth_link(post, 3)
)

# Calculate deltas for WET, NIW, and WL conditions
delta.wsas.wet <- expected[[1]][, 1] - expected[[1]][, 5]
delta.wsas.ew <- expected[[2]][, 1] - expected[[2]][, 5]
delta.wsas.wl <- expected[[3]][, 1] - expected[[3]][, 5]
```

Functional impairment, assessed using the WSAS, showed significant improvement across all three interventions. 
At one-week follow-up, the estimated improvement in WSAS scores was `r format_estimate(delta.wsas.wet, estimand = 'M')` for WET, `r format_estimate(delta.wsas.ew, estimand = 'M')` for NIW, and `r format_estimate(delta.wsas.wl, estimand = 'M')` for the waiting list.
Both WET (`r format_estimate(delta.wsas.wet/sd.wsas.wet - delta.wsas.wl/sd.wsas.wl)`) and NIW (`r format_estimate(delta.wsas.ew/sd.wsas.ew - delta.wsas.wl/sd.wsas.wl)`) resulted in greater improvements in functional impairment compared to the waiting list at the one week followup.
However, no significant differences were observed between them (`r format_estimate(delta.wsas.wet/sd.wsas.wet - delta.wsas.ew/sd.wsas.ew)`).
These outcomes were kept at the 2 month followup.

## Treatment Processes

We now shift our focus to understanding the processes underlying change in WET.
Our second, exploratory hypothesis proposed that the identified mechanisms of change would improve during therapy.
Detailed results for WET are presented in Table 2.
Below the table, we highlight the significant findings in WET, alongside the corresponding effects observed in the NIW group.
Difference between the groups is reported in all places where they are significant.
This comparison aims to identify differential effects while maintaining a primary focus on WET, which is the central topic of this paper.
Full results for NIW are available in the supplementary materials.

```{r processes_tab, file=here('scripts','tables.processes.R'), echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
```

```{r processes, echo=FALSE, include=FALSE, cache=FALSE}
if (!exists('process_models')) source(here('scripts','run.processes.R'))
get_change = function(proc, condition = 'wet'){(proc[[condition]][,1]-proc[[condition]][,2])/if_else(condition == 'wet', proc$sd.wet, proc$sd.ew)}

d.dts = get_change(process_models[['dts']])
d.bcss_self_pos = get_change(process_models[['bcss_self_pos']])
d.ifs = get_change(process_models[['ifs']])
d.mcq_confidence = get_change(process_models[['mcq_confidence']])
d.mcq_danger = get_change(process_models[['mcq_danger']])

d.mcq_confidence.ew = get_change(process_models[['mcq_confidence']], condition = 'ew')
d.dts.ew = get_change(process_models[['dts']], condition = 'ew')
d.bcss_self_pos.ew = get_change(process_models[['bcss_self_pos']], condition = 'ew')
d.ifs.ew = get_change(process_models[['ifs']], condition = 'ew')
d.mcq_danger.ew = get_change(process_models[['mcq_danger']], condition = 'ew')
```

Distress tolerance (DTS) improved significantly in the WET group (`r format_estimate(-d.dts)`), while it somewhat deteriorated in the NIW group (`r format_estimate(-d.dts.ew)`).
The difference between the groups was statistically significant (`r format_estimate(d.dts.ew - d.dts)`), aligning with our expectations.
WET likely encouraged participants to tolerate distress through structured exposure, whereas NIW may have reinforced avoidance behaviors.

Positive self-schemas, as measured by the BCSS, increased significantly in WET (`r format_estimate(-d.bcss_self_pos)`), suggesting that the intervention fostered more adaptive self-beliefs.
In contrast, no significant change was found in the NIW group (`r format_estimate(-d.bcss_self_pos.ew)`).

Perceived impact of future events (IFES) decreased significantly in the WET group (`r format_estimate(d.ifs)`).
No significant change was observed in the NIW group (`r format_estimate(d.ifs.ew)`).

For metacognitions measured by the MCQ, both (lack of) cognitive confidence (`r format_estimate(d.mcq_confidence)`) and negative beliefs about uncontrollability and danger (`r format_estimate(d.mcq_danger)`) increased significantly in WET.
However, no significant changes were found in the NIW group for either (lack of) cognitive confidence (`r format_estimate(d.mcq_confidence.ew)`) or negative beliefs about uncontrollability and danger (`r format_estimate(d.mcq_danger.ew)`).

These findings suggest several potential mechanisms of change for WET, which we investigate in the upcoming section.
The other mechanisms did not improve and thus will not be analyzed.

## Mechanisms of Change

```{r mechanisms, echo=FALSE, include=FALSE, cache=FALSE}
source(here('scripts','run.mechanisms.R'))
r.dts = extract.samples(mechanism_models$dts)$Rho[,2,5] 
r.dts.ew = extract.samples(mechanism_models$dts.ew)$Rho[,2,5] 
r.bcss_self_pos = extract.samples(mechanism_models$bcss_self_pos)$Rho[,3,5] 
r.mcq_confidence = extract.samples(mechanism_models$mcq_confidence)$Rho[,3,5] 
r.mcq_danger = extract.samples(mechanism_models$mcq_danger)$Rho[,3,5] 
```

We examined each process affected by WET to assess whether changes in these processes corresponded with symptom change.
Detailed results, including specific changes in processes and symptom levels, can be found in Table 3.

The only process found both to respond to WET and to change alongside symptoms was distress tolerance (DTS).
The correlation between changes in DTS and symptoms during treatment was `r format_estimate(r.dts, estimand = '*r*')`.
Given that NIW actively *decreased* distress tolerance, we investigated the effect of change in distress tolerance on change in symptoms within NIW as well.
We found that in contrast with WET, in NIW an increase in distress tolerance was actually correlated with an increase in symptoms `r format_estimate(r.dts.ew, estimand = '*r*')`.
This suggests a differential process where in WET distress tolerance decreases anxiety whereas in NIW anxiety is decreased via distraction and avoiding stress.

No processes were significantly predictive of change at follow-up.
However, improvements in positive self-schema (`r format_estimate(r.bcss_self_pos, estimand = '*r*')`), cognitive confidence (`r format_estimate(r.mcq_confidence, estimand = '*r*')`), and a reduction in negative beliefs about uncontrollability and danger (`r format_estimate(r.mcq_danger, estimand = '*r*')`) all weakly predicted change at follow-up.

Thus, these four processes warrant additional research when investigating the mechanisms of WET.

```{r mechanisms.table, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE, file=here('scripts','tables.mechanisms.R')}
```

## Response to Core Threats

```{r state.anx.code, echo=FALSE, include=FALSE, cache=TRUE}
if (!exists('mechanism_models')) source(here('scripts','run.mechanisms.R'))
r.bsam.change.all = extract.samples(mechanism_models$bsam.change.all)$Rho[,2,5] 
r.bsam.change.all.ew = extract.samples(mechanism_models$bsam.change.all.ew)$Rho[,2,5] 
r.dts.bsam.wet = extract.samples(mechanism_models$dts.bsam.wet)$Rho[,2,5] 
```

Alongside global mechanisms and metacognitions, we explored how participants’ threat responses changed in relation to their core threat.
To investigate this, we examined state anxiety before and after each writing session using the BSAM.
Participants rated their state anxiety while vividly imagining their core threat script.
This approach allowed us to assess both within-session and between-session changes in anxiety levels (see Figure \@ref(fig:stateanx)).

(ref:stateanx) Trajectory of state anxiety as a function of time. Dashed lines are the trajectory of pre intervention measurements. Arrows are within session changes, pointing at the post intervention measure.

```{r stateanx, file=here('scripts','figures.state.anx.R'), echo=FALSE, message=FALSE, warning=FALSE, fig.cap='(ref:stateanx)', cache=TRUE}
```

Our findings revealed distinct differences between the groups.
Both began with similarly high levels of state anxiety.
During the first session, the NIW group experienced a significant reduction in anxiety, reflecting their instruction to distract themselves from the threat.
In contrast, the WET group’s anxiety levels remained high and stable, consistent with their instruction to focus on the core threat.

By the second session, both groups had established new patterns:
the NIW group maintained the reduced anxiety achieved after session one, while the WET group experienced a significant reduction in anxiety even below that of the NIW group.
This between-session reduction in state anxiety predicted improvements in trait anxiety (OASIS), in line with meta-analysis showing that between session habituation is correlated with treatment outcome [e.g., @ruppEmotionalProcessingTheory2017].
However, stronger correlations were found in the WET group (`r format_estimate(r.bsam.change.all, estimand = '*r*')`) than in the NIW group (`r format_estimate(r.bsam.change.all.ew, estimand = '*r*')`).
It is reasonable to hypothesize that the reduction in trait anxiety in response to the core threat is caused by decrease in DTS, as seen above.
However, no correlation was found between change in DTS and change in state anxiety (`r format_estimate(r.dts.bsam.wet, estimand = '*r*')`).

The distinct trajectory of state anxiety in the WET group suggests a unique mechanism of change.
Specifically, the larger decrease in state anxiety between sessions in the WET group may indicate deeper processing of core threats, which in turn is associated with more robust symptom improvements.

Furthermore, within-session patterns diverged between the groups.
For the NIW group, writing reduced state anxiety, while for the WET group, it increased state anxiety.
This result serves as a manipulation check and aligns with the interventions' instructions: WET participants were encouraged to actively engage with their fears, while NIW participants were guided to distract themselves from them.

Interestingly, there was evidence of a slight reduction in the magnitude of within-session anxiety increases over time in the WET group.
This trend suggests that participants may have become more accustomed to the task, potentially reflecting a process of habituation.
